
---
title: "Temporal Dynamics and Predictive Modelling of Mpox in Nigeria (2017–2025)"
author: "Lanre Peter Daodu et al."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document: default
---

# Introduction

This document contains the complete reproducible analytical workflow for modelling Mpox epidemiological time-series data in Nigeria (2017–2025) using ARIMA and ARIMAX models.

# Package Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(ISOweek)
library(forecast)
library(tseries)
library(ggplot2)
library(patchwork)
library(readxl)
```

# Data Import

```{r data-import}
mpox_data <- read_xlsx("Mpox-Data_2017-2025V2.xlsx")

mpox_data <- mpox_data %>%
  rename(
    Year = `Year`,
    EpiWeek = `Epi-Week`,
    Suspected = `Suspected Cases`,
    Confirmed = `Confirmed Cases`,
    Death = `Death`
  ) %>%
  mutate(across(c(Suspected, Confirmed, Death), ~replace_na(., 0)))
```

# Create Weekly Date Index

```{r weekly-date}
mpox_data <- mpox_data %>%
  mutate(
    EpiWeek = as.numeric(EpiWeek),
    iso_week = sprintf("%02d", EpiWeek),
    iso_string = paste0(Year, "-W", iso_week, "-1"),
    WeekStartDate = ISOweek2date(iso_string)
  ) %>%
  arrange(WeekStartDate)
```

# Create Time Series Objects

```{r ts-objects}
confirmed_ts <- ts(mpox_data$Confirmed, start = c(2017,1), frequency = 52)
suspected_ts <- ts(mpox_data$Suspected, start = c(2017,1), frequency = 52)
death_ts     <- ts(mpox_data$Death,     start = c(2017,1), frequency = 52)
```

# Stationarity Testing

```{r stationarity}
adf.test(confirmed_ts)
kpss.test(confirmed_ts)

diff_confirmed <- diff(confirmed_ts)
adf.test(diff_confirmed)
kpss.test(diff_confirmed)
```

# ARIMA Model Estimation

```{r arima-fit}
auto_confirmed <- auto.arima(confirmed_ts, seasonal = FALSE)
summary(auto_confirmed)
checkresiduals(auto_confirmed)
```

# Forecasting

```{r forecasting}
forecast_horizon <- 52
fc_confirmed <- forecast(auto_confirmed, h = forecast_horizon)

autoplot(fc_confirmed) +
  labs(title="52-Week Forecast of Confirmed Mpox Cases",
       x="Year", y="Cases")
```

# Rolling-Origin Cross-Validation

```{r cross-validation}
set.seed(123)

y <- as.numeric(confirmed_ts)
n <- length(y)
initial <- 260
h <- 1

origins <- seq(initial, n-h)
errors <- numeric(length(origins))

for (i in seq_along(origins)) {
  train <- ts(y[1:origins[i]], frequency=52)
  fit <- auto.arima(train, seasonal=FALSE)
  pred <- forecast(fit, h=h)$mean[1]
  errors[i] <- y[origins[i]+h] - pred
}

MAE  <- mean(abs(errors))
RMSE <- sqrt(mean(errors^2))

tibble(MAE=MAE, RMSE=RMSE)
```

# Rainfall Covariate (ARIMAX)

```{r rainfall-import}
rainfall_data <- read_xlsx("Rainfall_data2017_2025.xlsx")
colnames(rainfall_data) <- c("Date","rainfall_mm")

rainfall_data$Date <- as.Date(rainfall_data$Date)

rain_weekly <- rainfall_data %>%
  mutate(WeekStartDate = floor_date(Date,"week")) %>%
  group_by(WeekStartDate) %>%
  summarise(rainfall_mm = mean(rainfall_mm, na.rm=TRUE))

mpox_rain <- left_join(mpox_data, rain_weekly, by="WeekStartDate")
mpox_rain$rainfall_mm[is.na(mpox_rain$rainfall_mm)] <- 0

rain_ts <- ts(mpox_rain$rainfall_mm, start=c(2017,1), frequency=52)
```

```{r arimax}
fit_arimax <- auto.arima(confirmed_ts, xreg = rain_ts, seasonal=FALSE)
summary(fit_arimax)

fc_arimax <- forecast(fit_arimax,
                      xreg = tail(rain_ts, forecast_horizon),
                      h = forecast_horizon)

autoplot(fc_arimax) +
  labs(title="ARIMAX Forecast with Rainfall Covariate")
```

# Descriptive Statistics

```{r descriptives}
mpox_summary <- mpox_data %>%
  group_by(Year) %>%
  summarise(
    Total_Suspected = sum(Suspected),
    Total_Confirmed = sum(Confirmed),
    Total_Death     = sum(Death),
    CFR = ifelse(Total_Confirmed==0, NA,
                 (Total_Death/Total_Confirmed)*100)
  )

mpox_summary
```

# Session Information

```{r session-info}
sessionInfo()
```
